Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-07-14 ebayes_rnaseq_heterosis.tex
through discussion

With almost 30,000 genes in our data set, we can evaluate the distribution assumptions for the gene-specific parameters. Figure \ref{f:scatterplot} provides marginal and bivariate histograms for the gene-specific parameters.
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{estimates}}
\caption{Marginal and bivariate histograms for gene-specific parameter posterior meansfor the B73 $\times$ Mo17 maize experiment. For reference, the marginal histograms have a dashed line indicating the appropriate marginal probability density functions in the hierarchical model with the MAP estimates.}
\label{f:scatterplot}
\end{figure}

@@

% Estimates of heterosis to compare with Ji et. al.
In addition to providing estimates of heterosis probabilities and effect sizes, this real data analysis exposed violations of our normality assumptions, indicating room to improve the model and increase the power to detect gene expression heterosis. Some of our ongoing and future research will develop the details of such an extension.

>>>>

@@

Although our method appears to improve upon existing methods, we believe our approach can be improved by refining the hierarchical model for the gene-specific parameter distribution. Figure \ref{f:scatterplot} shows marginal and bivariate histograms for the empirical Bayes posterior means for the gene-specific parameters. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{estimates}}
\caption{Marginal and bivariate histograms for gene-specific parameter posterior meansfor the B73 $\times$ Mo17 maize experiment. For reference, the marginal histograms have a dashed line indicating the appropriate marginal probability density functions in the hierarchical model with the MAP estimates.}
\label{f:scatterplot}
\end{figure}
These figures show markedly departure from marginal model assumptions, e.g. normality assumptions for $\phi_g$ and $\psi_g$, and independence assumptions, e.g. $\phi_g$ versus $\psi_g$ and $\alpha_g$ versus $\delta_g$. The plot of $\phi_g$ versus $\psi_g$ shows a pattern where the mean overdispersion decreases as the mean expression level increases. The plot of $\alpha_g$ versus $\delta_g$ shows a rotated V pattern where $\delta_g$ appears to be equal to $|\alpha_g|$. This pattern is consistent with Mendel's Law of Dominance where the hybrid has mean expression equal to the parent with higher mean expression.

In addition to improving the hierarchical distribution, we believe better estimates of the parameters of this distribution, i.e. the hyperparameters, will also improve detection of gene-expression heterosis. Our current method, based on moment matching of essentially independently estimated gene-specific parameters provides consistent estimators as the number of replicates per variety increases. But typically the number of replicates per variety is quite small, in our data there are only four replicates per variety, and therefore assymptotic justifications appear deficient. There are a variety of hyperparameter estimation approaches to explore, e.g. expectation-maximization algorithms or fully Bayesian approaches.

Notwithstanding these improvements, we believe our approach is a computationally efficient method that can immediately aid scientists who are interested in identifying candidate genes to explore a genetic mechanism of heterosis.
